---
title: "HW 03"
author: "Amit Chawla"
format:
  html:
    embed-resources: true
toc: true
---

## Initial Setup

```{r}
#| label: Initial Setup

if (!require("pacman")) 
  install.packages("pacman")

pacman::p_load(tidyverse,
               janitor,
               colorspace,
               broom,
               fs,
               scales,
               ggthemes,
               ggrepel,
               patchwork,
               ggimage,
               jpeg,
               glue,
               grid,
               forcats)

# set theme for ggplot2
ggplot2::theme_set(ggplot2::theme_minimal(base_size = 14))

# set width of code output
options(width = 65)

# set figure parameters for knitr
knitr::opts_chunk$set(
  fig.width = 7, # 7" width
  fig.asp = 0.618, # the golden ratio
  fig.retina = 3, # dpi multiplier for displaying HTML output on retina
  fig.align = "center", # center align figures
  dpi = 300 # higher dpi, sharper image
)

```

## 1 - Du Bois challenge.

```{r}
#| label: Du Bois Visualization

# 1. Base data
du_bois_income <- read_csv("data/income.csv")

# 2. Pivot + cleanup with reversed category order
du_bois_long <- du_bois_income |>
  pivot_longer(cols = Rent:Other, names_to = "Category", values_to = "Expenditure") |>
  mutate(
    Category = fct_relevel(Category, "Rent", "Food", "Clothes", "Tax", "Other"),  # For legend order
    Class = fct_rev(fct_inorder(Class))
  ) |>
  arrange(Class, desc(Category)) |>  # Reverse category order within each class
  group_by(Class) |>
  mutate(
    CenterPos = cumsum(Expenditure) - 0.5 * Expenditure,  # Recalculate positions
    Label = paste0(Expenditure, "%"),
    LabelColor = ifelse(Category == "Rent", "white", "black")
  ) |>
  ungroup()

# 3. Colors (match reference)
colors <- c("Rent" = "black", "Food" = "purple", "Clothes" = "sienna1", "Tax" = "slategray1", "Other" = "snow2")

# 4. Background image
bg_path <- "images/du-bois-bg.jpg"
bg <- jpeg::readJPEG(bg_path)
bg_grob <- rasterGrob(bg, width = unit(1, "npc"), height = unit(1, "npc"), interpolate = TRUE)

# 5. Final Plot
ggplot(du_bois_long, aes(x = Class, y = Expenditure, fill = Category)) +
  annotation_custom(bg_grob, xmin = -Inf, xmax = Inf, ymin = -Inf, ymax = Inf) +
  geom_col(width = 0.5, color = "white") +
  geom_text(aes(y = CenterPos, label = Label, color = LabelColor),
            size = 3, show.legend = FALSE, family = "mono") +
  scale_color_identity() +
  scale_fill_manual(values = colors, breaks = c("Rent", "Food", "Clothes", "Tax", "Other")) +
  scale_x_discrete(labels = c("1,000     $1,125 \nAND OVER          ", "$750-1000   $880   ", "$500-750   $547   ", "$400-500   $433.82", "$300-400   $335.66", "$200-300   $249.45", "$100-200   $139.10")) +
  coord_flip(clip = "off") +
  labs(
    x = NULL, y = NULL,
    title = "INCOME AND EXPENDITURE OF 150 NEGRO FAMILIES IN ATLANTA, GA., U.S.A."
  ) +
  theme_minimal(base_family = "mono") +
  theme(
    legend.position = "top",
    plot.title = element_text(face = "bold", size = 11, hjust = 0.5),
    panel.grid = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_text(size = 7),
    text = element_text(size = 9),
    legend.title = element_blank(),
    legend.text = element_text(size = 9),
    legend.key.size = unit(0.3, "cm")
  ) +
  annotate(
    geom = "text",
    x = "$100-200",
    y = -0,
    label = "CLASS    ACTUAL AVERAGE                                                                                  \n\n\n",
    size = 2,
    color = "black",
    family = "mono",
    hjust = 0
  )

```

## 2 - COVID survey - interpret

### Interpretation of the COVID-19 Vaccine Attitudes Visualization

This visualization is packed with information about how medical and nursing students across the U.S. feel about the COVID-19 vaccine. It’s a grid of facets where each column represents a different statement about the vaccine, rated on a Likert scale (1 = Strongly Agree to 5 = Strongly Disagree), and each row breaks down the responses by demographic factors like age, gender, race, profession, and vaccination status. The points show the mean score for each group, and the error bars stretch from the 10th to 90th percentiles, giving a sense of how much opinions vary. The first row, labeled "All," shows the overall responses without splitting by any demographic factors.

#### Overall Observations

Starting with the "All" row, it’s cool to see that most students lean toward positive views about the vaccine. For statements like "I believe the vaccine is safe," "Getting the vaccine will make me feel safer at work," and "I will recommend the vaccine to others," the mean scores hover around 1.5 to 2, which means they’re generally agreeing—either strongly or somewhat. That feels right to me since these are future healthcare pros who probably trust science. But then, for "I am concerned about the safety and side effects of the vaccine," the mean jumps to about 3 (neutral), and the error bars go from like 1.5 to 4.5. So, even though they trust the vaccine, a lot of them are still worried about side effects, which I get because it was developed so fast.

#### Example 1: Asian Respondents’ Mixed Feelings

One thing that caught my eye was in the "Race" category with Asian students. For "I believe the vaccine is safe," the error bars are huge—like, they go from 1 to 5! That means some strongly agree it’s safe, while others totally disagree, which is wild variation. I didn’t expect that much difference within one group. But then, for "I will recommend the vaccine to others," the mean is around 2, and the error bars are way tighter, maybe 1 to 3. That’s weird to me—if you’re all over the place on safety, I’d think you’d be unsure about recommending it too. Maybe they feel they *should* promote it, even if they’re not totally sold on it themselves, like it’s their duty or something during a pandemic.

#### Example 2: Nursing vs. Medical Students

Looking at the "Profession" row, I noticed nursing and medical students don’t line up as much as I thought they would. For stuff like "I trust the information I have received," nursing students have tight error bars, maybe 1 to 2.5, so they’re pretty consistent. Medical students, though, have wider ones, like 1 to 4, showing more mixed opinions. I figured both groups would think alike since they’re both in healthcare and learning the same science. This makes me wonder if medical students are more skeptical because they’re digging into research more, while nursing students might just go with what they’re taught. It’s not what I expected!

#### Example 3: Vaccinated Students Still Worried

In the "Had COVID vaccine" row, students who said "Yes" to being vaccinated have means around 1 to 1.5 for positive statements like "I believe the vaccine is safe," which makes sense—if you got it, you probably trust it. But for "I am concerned about side effects," the mean is still around 3, with error bars from 1.5 to 4.5. That’s interesting because I thought if you’re vaccinated, you’d be less worried. It shows even people who took it aren’t totally chill about risks, which fits with how some might’ve gotten it for school or work but still have doubts about a new vaccine.

#### Wrapping Up

This plot shows that medical and nursing students mostly trust the COVID-19 vaccine, but they’re not all-in—side effect worries are real across the board. The differences between groups, like race or profession, add layers I didn’t expect, making it clear that even future doctors and nurses don’t all see it the same way. It’s pretty fascinating how complicated their views are!

## 3 - COVID survey - reconstruct

```{r}
#| label: Loading data for Covid survey visualization

# Loading Data

covid_check <- read.csv("data/covid-survey.csv")
#head(covid)

# After watching the head of the data set we need to remove the first row which just says this - 'likert_survey' And doesn't have actual data

covid <- read.csv("data/covid-survey.csv", skip = 1)

dim(covid)
```

```{r}
#| label: Here sipping Unnecessary Starting Row from Data set

# I would have add [na = c(".", "")] during loading data but just to follow along the step doing it separatly below.
covid_filtered_na <- covid |>
  mutate(across(-response_id, ~ na_if(trimws(.), ""))) |>
  mutate(across(-response_id, ~ na_if(., "."))) |>
  filter(if_any(-response_id, ~ !is.na(.)))

dim(covid_filtered_na)


```

```{r}
#| label: Adding appropriate labels to the explantory variables

# Updating Labels
covid_relabeled <- covid_filtered_na |>
  mutate(
    exp_already_vax = recode(exp_already_vax, `0` = "No", `1` = "Yes"),
    exp_flu_vax = recode(exp_flu_vax, `0` = "No", `1` = "Yes"),
    exp_profession = recode(exp_profession, `0` = "Medical", `1` = "Nursing"),
    exp_gender = recode(exp_gender,
                        `0` = "Male",
                        `1` = "Female",
                        `3` = "Non-binary third gender",
                        `4` = "Prefer not to say"),
    exp_race = recode(exp_race,
                      `1` = "American Indian / Alaskan Native",
                      `2` = "Asian",
                      `3` = "Black / African American",
                      `4` = "Native Hawaiian / Other Pacific Islander",
                      `5` = "White"),
    exp_ethnicity = recode(exp_ethnicity,
                           `1` = "Hispanic / Latino",
                           `2` = "Non-Hispanic / Non-Latino"),
    exp_age_bin = recode(exp_age_bin,
                         `0` = "<20",
                         `20` = "21–25",
                         `25` = "26–30",
                         `30` = ">30")
  )

dim(covid_relabeled)


```

```{r}
#| label: Pivoting the columns

covid_survey_longer <- covid_relabeled |>

# This pivot longer combines all explanatory columns and their values into two columns. Pivoting multiple columns of data into two columns to be able to address easily while plotting.

  pivot_longer(
    cols = starts_with("exp_"),
    names_to = "explanatory",
    values_to = "explanatory_value"
  ) |>
  filter(!is.na(explanatory_value)) |>

# This pivot longer combines all response value columns into and their values into two columns. Similar to the above one it will also help in plotting the data.

  pivot_longer(
    cols = starts_with("resp_"),
    names_to = "response",
    values_to = "response_value"
  )

covid_survey_longer

covid_survey_longer |>
  distinct(explanatory_value)
```

```{r}
#| label: Calculating Summary Statistics by Group

covid_survey_longer <- covid_survey_longer |>
  mutate(response_value = as.numeric(response_value))

covid_survey_summary_stats_by_group <- covid_survey_longer |>
  group_by(explanatory, explanatory_value, response) |>
  summarise(
    mean = mean(response_value, na.rm = TRUE),
    low = quantile(response_value, 0.1, na.rm = TRUE),
    high = quantile(response_value, 0.9, na.rm = TRUE),
    .groups = "drop"
  )

covid_survey_summary_stats_by_group



```

```{r}
#| label: Summarizing Stats and binding rows for Visualization

covid_survey_summary_stats_all <- covid_survey_longer |>
  group_by(response) |>
  summarise(
    mean = mean(response_value, na.rm = TRUE),
    low = quantile(response_value, 0.1, na.rm = TRUE),
    high = quantile(response_value, 0.9, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(
    explanatory = "All",
    explanatory_value = ""
  )

covid_survey_summary_stats_all

# checking distint values to see if all variable values are coming
covid_survey_summary_stats_all |>
  distinct(explanatory_value)

# Binding the rows

covid_survey_summary_stats <- bind_rows(
  covid_survey_summary_stats_all,
  covid_survey_summary_stats_by_group
)


covid_survey_summary_stats


```

```{r}
#| label: Relevels factors, Set Labels, and Create Plot

# Relevel factors
covid_survey_summary_stats <- covid_survey_summary_stats |>
  mutate(
    explanatory_value = fct_relevel(explanatory_value, ">30", "26-30", "21-25", "<20"),
    explanatory = fct_relevel(explanatory, "All", "exp_age_bin", "exp_gender", "exp_race", "exp_ethnicity", "exp_profession", "exp_already_vax", "exp_flu_vax"),
    explanatory_value = fct_relevel(explanatory_value, "Male", "Female", "Non-binary third gender", "Prefer not to say"),
    explanatory_value = fct_relevel(explanatory_value, "American Indian/Alaskan Native", "Asian", "Black/African American", "Native Hawaiian/Other Pacific Islander", "White"),
    explanatory_value = fct_relevel(explanatory_value, "Non-Hispanic/Non-Latino", "Hispanic/Latino"),
    explanatory_value = fct_relevel(explanatory_value, "Medical", "Nursing"),
    explanatory_value = fct_relevel(explanatory_value, "No", "Yes")
  )

# Set labels
covid_survey_summary_stats <- covid_survey_summary_stats |>
  mutate(
    explanatory = recode(explanatory, "exp_age_bin" = 'Age', "exp_gender" = 'Gender', "exp_race" = 'Race', "exp_ethnicity" = 'Ethnicity', "exp_profession" = 'Profession', "exp_already_vax" = 'Had COVID vaccine', "exp_flu_vax" = 'Had flu vaccine this year'),
    response = fct_relevel(response, "resp_safety", "resp_feel_safe_at_work", "resp_concern_safety", "resp_confidence_science",   "resp_trust_info", "resp_will_recommend"),
    response = recode(response,
                      "resp_safety" = 'Based on my understanding, I believe the vaccine is safe',
                      "resp_confidence_science" = 'I am confident in the scientific vetting process for the new COVID vaccines',
                      "resp_feel_safe_at_work" = 'Getting the vaccine will make me feel safer at work',
                      "resp_will_recommend" = 'I will recommend the vaccine to family, friends, and community members',
                      "resp_trust_info" = 'I trust the information that I have received about the vaccines',
                      "resp_concern_safety" = 'I am concerned about the safety and side effects of the vaccine')
  )

# Creating the plot
covid_survey_summary_stats |>
  ggplot(aes(x = mean, y = factor(explanatory_value))) +
  geom_point(size = 0.75) +
  geom_errorbarh(aes(xmin = low, xmax = high), height = 0.3) +
  facet_grid(
    rows = vars(explanatory),
    cols = vars(response),
    scales = "free_y",
    space = "free_y",
    labeller = labeller(explanatory = label_wrap_gen(15), response = label_wrap_gen(15))
  ) +
  scale_x_continuous(breaks = 1:5) +
  labs(
    x = "Mean Likert score\n(Error bars range from 10th to 90th percentile)",
    y = NULL,
  ) +
  theme(
    strip.text = element_text(size = 6),
    strip.text.y = element_text(angle = 0),
    axis.text.y = element_text(size = 6),
    axis.text.x = element_text(size = 6),
    panel.spacing = unit(0, "lines"),
    panel.spacing.x = unit(0.3, "lines"),
    axis.title.x = element_text(size = 8),
    panel.grid = element_blank(),
    strip.background = element_rect(fill = "gray90", color = "black")
  )
```

## 4 - COVID survey - re-reconstruct

```{r}

# Calculating summary statistics for identity groups
covid_survey_summary_stats_by_group_2 <- covid_survey_longer |>
  group_by(explanatory, explanatory_value, response) |>
  summarise(
    mean = mean(response_value, na.rm = TRUE),
    low = quantile(response_value, 0.25, na.rm = TRUE),
    high = quantile(response_value, 0.75, na.rm = TRUE),
    .groups = "drop"
  )

# Calculating summary statistics for all
covid_survey_summary_stats_all_2 <- covid_survey_longer |>
  group_by(response) |>
  summarise(
    mean = mean(response_value, na.rm = TRUE),
    low = quantile(response_value, 0.25, na.rm = TRUE),
    high = quantile(response_value, 0.75, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(
    explanatory = "All",
    explanatory_value = ""
  )

# Binding the rows
covid_survey_summary_stats_2 <- bind_rows(
  covid_survey_summary_stats_all_2,
  covid_survey_summary_stats_by_group_2
)

# Relevels factors, Set Labels, and Create Plot

# Relevel factors
covid_survey_summary_stats_2 <- covid_survey_summary_stats_2 |>
  mutate(
    explanatory_value = fct_relevel(explanatory_value, ">30", "26-30", "21-25", "<20"),
    explanatory = fct_relevel(explanatory, "All", "exp_age_bin", "exp_gender", "exp_race", "exp_ethnicity", "exp_profession", "exp_already_vax", "exp_flu_vax"),
    explanatory_value = fct_relevel(explanatory_value, "Male", "Female", "Non-binary third gender", "Prefer not to say"),
    explanatory_value = fct_relevel(explanatory_value, "American Indian/Alaskan Native", "Asian", "Black/African American", "Native Hawaiian/Other Pacific Islander", "White"),
    explanatory_value = fct_relevel(explanatory_value, "Non-Hispanic/Non-Latino", "Hispanic/Latino"),
    explanatory_value = fct_relevel(explanatory_value, "Medical", "Nursing"),
    explanatory_value = fct_relevel(explanatory_value, "No", "Yes")
  )

# Set labels
covid_survey_summary_stats_2 <- covid_survey_summary_stats_2 |>
  mutate(
    explanatory = recode(explanatory, "exp_age_bin" = 'Age', "exp_gender" = 'Gender', "exp_race" = 'Race', "exp_ethnicity" = 'Ethnicity', "exp_profession" = 'Profession', "exp_already_vax" = 'Had COVID vaccine', "exp_flu_vax" = 'Had flu vaccine this year'),
    response = fct_relevel(response, "resp_safety", "resp_feel_safe_at_work", "resp_concern_safety", "resp_confidence_science",   "resp_trust_info", "resp_will_recommend"),
    response = recode(response,
                      "resp_safety" = 'Based on my understanding, I believe the vaccine is safe',
                      "resp_confidence_science" = 'I am confident in the scientific vetting process for the new COVID vaccines',
                      "resp_feel_safe_at_work" = 'Getting the vaccine will make me feel safer at work',
                      "resp_will_recommend" = 'I will recommend the vaccine to family, friends, and community members',
                      "resp_trust_info" = 'I trust the information that I have received about the vaccines',
                      "resp_concern_safety" = 'I am concerned about the safety and side effects of the vaccine')
  )

# Creating the plot
covid_survey_summary_stats_2 |>
  ggplot(aes(x = mean, y = factor(explanatory_value))) +
  geom_point(size = 0.75) +
  geom_errorbarh(aes(xmin = low, xmax = high), height = 0.5) +
  facet_grid(
    rows = vars(explanatory),
    cols = vars(response),
    scales = "free_y",
    space = "free_y",
    labeller = labeller(explanatory = label_wrap_gen(15), response = label_wrap_gen(15))
  ) +
  scale_x_continuous(breaks = 1:5) +
  labs(
    x = "Mean Likert score\n(Error bars range from 25th to 75th percentile)",
    y = NULL,
  ) +
  theme(
    strip.text = element_text(size = 6),
    strip.text.y = element_text(angle = 0),
    axis.text.y = element_text(size = 6),
    axis.text.x = element_text(size = 6),
    panel.spacing = unit(0, "lines"),
    panel.spacing.x = unit(0.3, "lines"),
    axis.title.x = element_text(size = 8),
    panel.grid = element_blank(),
    strip.background = element_rect(fill = "gray90", color = "black")
  )

```

\
This plot differs quite a bit from the one in question 3 because the error bars now reflect the 25th to 75th percentiles (interquartile range) instead of the 10th to 90th percentiles, shrinking the range by 50 percentiles overall. This change makes the error bars much tighter around the mean Likert scores, giving a clearer picture of where most responses cluster rather than the full spread. It’s especially noticeable in columns like “I believe the vaccine is safe” and “I am concerned about side effects,” where the error bars are now more concentrated, even when broken down by factors like age or race. Comparing this to the previous plot feels like switching from a wide scatter to something closer to a box plot, showing the middle 50% of responses more clearly.

One cool thing that stands out is that non-binary third gender respondents’ error bar for “I believe the vaccine is safe” looks pretty similar to before, still hovering around a neutral mean with a tight range. This might suggest a small sample size, which could explain the lack of change. Another new observation is that for “Had COVID vaccine” (Yes vs. No), the “Yes” group shows a tight error bar around 1-2 for “I will recommend the vaccine,” indicating strong agreement, while the “No” group spreads wider toward 3-4, showing more uncertainty. I didn’t expect that big a difference—maybe those who didn’t get it are less confident in pushing it on others.

Overall, the shorter error bars suggest the data is more clustered between the 25th and 75th percentiles, which aligns with the Central Limit Theorem—larger sample sizes (likely over 30 here) tend to push distributions toward normal. This tighter focus helps me see that most medical and nursing students are pretty consistent in their views, especially on trust and recommendation, even if some worry about side effects.

## 5 - COVID survey - another view
